---
title: "Whale Shark Movement Analysis"
author: "Madelyne Zhang, Leo Lyu"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: pygments
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}
library(tidyverse)
library(lubridate)
library(sf)
library(patchwork)
#install.packages("remotes")
#install.packages(
#  "aniMotum",
#  repos = c(
#    "https://cloud.r-project.org",
#    "https://ianjonsen.r-universe.dev"
#  ),
#  dependencies = TRUE
#)

library(aniMotum)
```

### Data

```{r}
sharks <- read_csv("final_whale_shark.csv")

sharks %>% 
  count(id)

sharks <- sharks %>%
  mutate(
    date = stringr::str_replace(date, "-\\s+", "-"),
    date = ymd_hms(date)
  )

```

```{r}
# Check number of observations per individual 
sharks %>% 
    count(id) %>% 
    arrange(desc(n))
```

```{r}
# Summary of Argos location classes
sharks %>% count(lc)
```

```{r}
sharks <- sharks %>%  
  arrange(id,date)
sharks <- sharks %>%
  filter(!is.na(date))

# Largest time gap data check 
sharks %>% 
  filter (id == "172246")
sharks %>% 
  filter (id == "157790")

sharks %>% 
  filter (id =="203637")
sharks %>% 
  filter (id =="203638")
sharks %>% 
  filter (id =="203639")
sharks %>% 
  filter (id =="203644")
sharks %>% 
  filter (id =="203646")
```


```{r}
# Split for individuals with large temporal gaps - 172246 and 157790
# Also with 5 sharks that have huge movement that is "across the continent" after the MPM interpolation, which doesn't make sense biologically, so I headed back here at first to split these 5 sharks observations in advance. 

split_table <- tibble(
  id = c(
    "172246",
    "157790",
    "203637",
    "203638",
    "203639",
    "203644",
    "203646"
  ),
  split_time = ymd_hms(c(
    "2017-09-30 16:21:00",
    "2018-03-28 22:30:00",
    "2020-06-13 03:50:58",
    "2020-06-14 01:25:45",
    "2020-06-13 03:52:39",
    "2020-06-13 03:52:00",
    "2019-09-08 17:44:00"
    
  ))
)


```

### EDA

#### Observation time gap between sharks

```{r}
# Compute time gaps between consecutive observations within each individual
sharks_dt <- sharks %>%
  arrange(id, date) %>%
  group_by(id) %>%
  mutate(
    dt_hours = as.numeric(difftime(date, lag(date), units = "hours"))
  ) %>%
  ungroup()

# Time gap statistics per individual
sharks_dt %>%
  group_by(id) %>%
  summarise(
    n_obs = n(),
    mean_dt = mean(dt_hours, na.rm = TRUE),
    median_dt = median(dt_hours, na.rm = TRUE),
    max_dt = max(dt_hours, na.rm = TRUE)
  ) %>%
  arrange(desc(max_dt))

```

```{r}
sharks_dt %>%
  summarise(
    min_hours = min(dt_hours, na.rm = TRUE),
    q1_hours = quantile(dt_hours, 0.25, na.rm = TRUE),
    median_hours = median(dt_hours, na.rm = TRUE),
    mean_hours = mean(dt_hours, na.rm = TRUE),
    q3_hours = quantile(dt_hours, 0.75, na.rm = TRUE),
    max_hours = max(dt_hours, na.rm = TRUE)
  )
```

The median of the consecutive time gap is 0.73 hours, but the maximum gap is around 2.9 years, where the tracking window has a huge gap.
Raw Argos telemetry exhibited highly irregular temporal sampling, with frequent short-interval transmissions interspersed with multi-month to multi-year gaps.

```{r}
# Time gap histogram 
ggplot(sharks_dt, aes(x = dt_hours)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  theme_bw() +
  labs(
    x = "Time gap between observations (hours, log scale)",
    y = "Count"
  )

```


```{r}
sharks <- sharks %>%
  mutate(id = as.character(id))

sharks <- sharks %>%
  left_join(split_table, by = "id") %>%
  mutate(
    id = case_when(
      is.na(split_time) ~ id,
      date <= split_time ~ paste0(id, "_a"),
      date >  split_time ~ paste0(id, "_b")
    )
  ) %>%
  select(-split_time)

sharks %>% 
  count(id)
```
Initially, the raw df contains 60 sharks, but since two of them have long time gap which caused non-convergence in the first attempt of MPM, and 5 of the sharks' movement track contains the "over-continent" track, which doesn't make sense biologically, so these sharks are split into _a and _b to make further analysis. Therefore, with 7 sharks split to _a and _b, 67 sharks are remained. 

```{r}
# Raw Spatial tracks for each individual(points)
ggplot(sharks, aes(x = lon, y = lat)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~id) +
  theme_bw()
```

```{r}
# Raw Spatial tracks for each individual(path)
ggplot(sharks, aes(x = lon, y = lat)) +
  geom_path(alpha = 0.4) +
  facet_wrap(~id) +
  theme_bw()

```

```{r}
# Check for id = 203637
# Geom_point
p1 <- ggplot() +
  theme_bw() +
  geom_point(
    data = sharks %>% dplyr::filter(id == "203637_a"),
    aes(x = lon, y = lat)
  )

print(p1)

# Geom_path
p2 <- ggplot() +
  theme_bw() +
  geom_path(
    data = sharks %>% dplyr::filter(id == "203637_a"),
    aes(x = lon, y = lat)
  )

print(p2)
```

#### Velocity check

```{r}
# Stepwise speed between consecutive observations
sharks_speed <- sharks_dt %>%
  mutate(
    dist_m = geosphere::distHaversine(
      cbind(lon, lat),
      cbind(lag(lon), lag(lat))
    ),
    speed_ms = dist_m / (dt_hours * 3600)
  )
```

```{r}
ggplot(sharks_speed, aes(x = speed_ms)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  theme_bw() +
  labs(x = "Raw speed (m/s, log scale)")
```

```{r}
# proportion of raw steps exceeding 2.7m/s
sharks_speed %>%
  summarise(
    prop_over_2_7 = mean(speed_ms > 2.7, na.rm = TRUE)
  )
```

Approximately 12% of observed movement steps exceeded a biologically informed maximum swimming speed of 2.7 m/s.

#### Sanity check

Since the SSM model needs at least 3 points to fit, there are 9 sharks with less than 3 observations, which will be excluded from our further analysis.

```{r}
# Exclude individuals with fewer than three observations 
sharks %>%
  count(id) %>%
  filter(n < 3)

sharks_ssm <- sharks %>%
  group_by(id) %>%
  filter(n() >= 3) %>%
  ungroup()

sharks_ssm %>% 
  count(id)
```

58 sharks remaining in the sharks_ssm dataframe.

```{r}
# Sanity check 

# whether date contains invalid data 
sharks_ssm %>%
  summarise(
    n_na_date = sum(is.na(date)),
    n_inf_date = sum(!is.finite(as.numeric(date)))
  )

# Check whether lon / lat contains invalid data
sharks_ssm %>%
  summarise(
    n_na_lon = sum(is.na(lon)),
    n_na_lat = sum(is.na(lat))
  )

# Check whether time is monotonically increasing 
sharks_ssm %>%
  group_by(id) %>%
  summarise(
    any_non_increasing_time = any(diff(date) <= 0, na.rm = TRUE)
  ) %>%
  filter(any_non_increasing_time)

```

For the sharks that have >= 3 observations, there are 0 observations of NA date as they were initially filted out. 
There are also non_increasing_time observations, which will be excluded.

```{r}
# Check problem ids observations
problem_ids <- sharks_ssm %>%
  group_by(id) %>%
  summarise(any_non_increasing_time = any(diff(date) <= 0, na.rm = TRUE)) %>%
  filter(any_non_increasing_time) %>%
  pull(id)
```

```{r}
# Duplicated rows
sharks_ssm %>%
  filter(id %in% problem_ids) %>%
  count(id, date) %>%
  filter(n > 1) %>%
  arrange(desc(n))
```

1 duplicated entries found, will be removed.

#### Cleaning dataset for the SSM

```{r}
# Final cleaned dataset used for state-space modelling

sharks_ssm_clean <- sharks %>%
  # remove rows with invalid time or coordinates
  filter(
    !is.na(date),
    is.finite(as.numeric(date)),
    !is.na(lon),
    !is.na(lat)
  ) %>%
  
  # order observations within each individual
  arrange(id, date) %>%
  
  # remove duplicated timestamps within individuals
  distinct(id, date, .keep_all = TRUE) %>%
  
  # ensure each individual has enough data and strictly increasing time
  group_by(id) %>%
  filter(
    n() >= 3,
    all(diff(date) > 0)
  ) %>%
  ungroup()
```

In sharks_ssm_clean, all 58 sharks have at least three valid observations with valid timestamps and locations.

```{r}
# Sanity check for clean data 
nrow(sharks)
nrow(sharks_ssm_clean)

length(unique(sharks$id))
length(unique(sharks_ssm_clean$id))

sharks_ssm_clean %>%
  group_by(id) %>%
  summarise(any_bad_time = any(diff(date) <= 0)) %>%
  filter(any_bad_time)
```

There are 58 sharks (12142 entries) remaining.

### SSM Model Fit

```{r}
# State-space model settings:
# - model: correlated random walk (crw)
# - time.step: 24 hours (daily regularization)
# - vmax: 2.7 m/s (biologically informed maximum swimming speed)

sharks_crw_ssm <- fit_ssm(
  x = sharks_ssm_clean,
  model = "crw",
  time.step = 24,     
  vmax = 2.7,          
  control = ssm_control(
    verbose = 0       
  )
)

# Extract fitted locations (at observation times)
sharks_fitted <- grab(sharks_crw_ssm, what = "fitted") %>%
  as_tibble()

# Extract predicted locations (regular 24 h interval)
sharks_predicted <- grab(sharks_crw_ssm, what = "predicted") %>%
  as_tibble()

```

```{r}
# Dataframe with convergence indicators 
sharks_crw_ssm
```

```{r}
# Identify individuals with non-converged SSM fits
nonconv_ids <- sharks_crw_ssm %>%
filter(!converged | !pdHess) %>%
pull(id)

nonconv_ids
```

One individuals ("172246_b") did not converge.

```{r}
# Inspect observation histories of non-converged individuals
sharks_ssm_clean %>% 
  filter(id == "172246_b")
```

2017 jumped to 2020, which is probably the reason of non-convergence.


```{r}
# Retain SSM fits that converged with a positive-definite Hessian
ssm_valid <- sharks_crw_ssm %>%
filter(converged, pdHess)

# Record IDs excluded due to non-convergence
ssm_excluded_ids <- sharks_crw_ssm %>%
filter(!converged | !pdHess) %>%
pull(id)

ssm_excluded_ids
```

After excluding non-converged fits, 57 sharks remained for further analysis.

```{r}
# Extract fitted and predicted locations from valid SSM fits only
sharks_fitted <- grab(ssm_valid, what = "fitted") %>%
as_tibble()

sharks_predicted <- grab(ssm_valid, what = "predicted") %>%
as_tibble()
```

#### Visualization of model fit

To illustrate model behavior under different data conditions, predicted tracks were visualized for a small number of representative individuals.

```{r}
# Individual with the minimum number of observations required for SSM fitting 
three_obs_id <- "132071"

plot(
ssm_valid |> dplyr::filter(id == three_obs_id),
what = "predicted",
type = 1
)

plot(
ssm_valid |> dplyr::filter(id == three_obs_id),
what = "predicted",
type = 2
)

```

Individuals with extremely sparse observations showed high positional uncertainty and were not used for behavioural interpretation.

```{r}
# Individual with one of the largest temporal gaps between observations
large_gap_id <- "112557"

plot(
ssm_valid |> dplyr::filter(id == large_gap_id),
what = "predicted",
type = 1
)

plot(
ssm_valid |> dplyr::filter(id == large_gap_id),
what = "predicted",
type = 2
)

```

```{r}
sharks_ssm_clean %>% 
  filter(id == "112557")
```

For this shark (id =112557), the shark has only 54 observations in 2012, with a gap of almost one year, followed by the observation almost one year, where the one year in between have huge uncertainties according to the graph.

```{r}
# Individual with a more typical observation pattern
typical_id <- "164968"

plot(
ssm_valid |> dplyr::filter(id == typical_id),
what = "predicted",
type = 1
)

plot(
ssm_valid |> dplyr::filter(id == typical_id),
what = "predicted",
type = 2
)
```

This individual shows a more typical observation pattern, with stable reconstructed tracks and moderate uncertainty.

```{r}
# Typical 
ty_id <- "203645"

plot(
ssm_valid |> dplyr::filter(id == ty_id),
what = "predicted",
type = 1
)

plot(
ssm_valid |> dplyr::filter(id == ty_id),
what = "predicted",
type = 2
)
```

Visualization shark track on a spatial shapefile:

```{r}
# Visualize predicted track for a representative individual
aniMotum::map(
ssm_valid |> dplyr::filter(id == "164968"),
what = "predicted"
)
# Compared with initial raw data
raw_164968 <- ggplot() +
  theme_bw() +
  geom_point(
    data = sharks %>% dplyr::filter(id == "164968"),
    aes(x = lon, y = lat)
  )

print(raw_164968)

```

All 57 shark tracks were visualized on a spatial shapefile:

```{r}
# Visualize predicted tracks for all converged individuals
aniMotum::map(
ssm_valid,
what = "predicted"
)
```

### SSM diagnostics

The diagnostics are applied to the above three sharks, with 1.
time series plot 2.
QQ plot 3.
ACF

```{r}
# SSM diagnostics for representative individuals
# - sparse observations
# - large temporal gaps
# - typical observation pattern

diagnostic_ids <- c("132071", "112557", "164968")

ssm_diag <- ssm_valid |>
dplyr::filter(id %in% diagnostic_ids)

# OSAR (one step ahead residuals)
res_diag <- osar(ssm_diag)

# Time series residuals 
plot(
res_diag,
type = "ts",
pages = 1
)

# QQ plot 
plot(
res_diag,
type = "qq",
pages = 1
)

#ACF
plot(
res_diag,
type = "acf",
pages = 1
)

```

Overall, the diagnostic plots do not raise major concerns about model fit.
Residuals show no clear trends over time, their distribution is broadly consistent with model assumptions, and there is little remaining autocorrelation.
This suggests the fitted state-space model adequately captures the movement dynamics in the data.

Note: Since 175953 and 203646_b failed to converge in MPM in later steps, their diagnostics are checked

```{r}
# Additional SSM diagnostics for individuals that later failed MPM convergence
diagnostic_ids_MPM_failed <- c("175953", "203646_b")

ssm_diag_MPM_failed <- ssm_valid |>
dplyr::filter(id %in% diagnostic_ids_MPM_failed)

# OSAR (One-step-ahead residuals)
res_diag_MPM_failed <- osar(ssm_diag_MPM_failed)

# Time series residuals 
plot(
res_diag_MPM_failed,
type = "ts",
pages = 1
)

# QQ plot 
plot(
res_diag_MPM_failed,
type = "qq",
pages = 1
)

#ACF
plot(
res_diag_MPM_failed,
type = "acf",
pages = 1
)
```

For the shark with id = 175953, the maximum time gap between consecutive observations is 2313.52 hours.
Despite this long gap, the SSM diagnostics do not show clear violations of model assumptions.
The residuals are generally centered around zero without strong temporal trends, the QQ plots show only mild tail deviations, and the ACF plots for both x and y mostly fall within the confidence bounds, indicating that temporal dependence is adequately captured by the CRW model.

For the shark with id = 203646_b, the SSM diagnostics for this individual do not show major violations of model assumptions.
Residuals are centred around zero with no strong temporal trends, their distribution broadly follows model expectations, and there is little remaining autocorrelation in either spatial dimension.
These results suggest that the CRW-based state-space model provides an adequate description of the movement process for this track.

### MPM (Move Persistence Model)

```{r}
# Prepare input data for MPM
# Retain individuals with at least 20 predicted locations
pmm_data <- sharks_predicted %>%
dplyr::group_by(id) %>%
dplyr::filter(dplyr::n() >= 20) %>%
dplyr::ungroup()

# Fit move persistence model
pmm_fit <- fit_mpm(
pmm_data,
model = "mpm",
control = mpm_control(verbose = 0)
)

# MPM varying gamma visulization for selected sharks
rep_ids_gamma <- c("184025", "139119", "184029", "220359")

pmm_data_sub <- sharks_predicted %>%
  dplyr::filter(id %in% rep_ids_gamma) %>%
  dplyr::group_by(id) %>%
  dplyr::filter(dplyr::n() >= 20) %>%
  dplyr::ungroup()

pmm_fit_sub <- fit_mpm(
  pmm_data_sub,
  model = "mpm",
  control = mpm_control(verbose = 0)
)

p_mpm <- plot(pmm_fit_sub, pages = 1)
print(p_mpm)

# Save plot
ggsave(
  filename = "g_over_time_representative_sharks.png",
  plot = p_mpm,
  width = 8,
  height = 6,
  dpi = 300
)
```


```{r}
# Dataframe with indicators of convergence of MPM
pmm_fit
```

After excluding individuals with fewer than 20 predicted locations, 33 sharks were retained for MPM fitting.
Among these, the model did not converge for two individuals (175953 and 203646_b), which were excluded from subsequent analyses.

### Data Extraction and preparation

```{r}
# Extract fitted move persistence estimates
mpm_gamma <- grab(pmm_fit, what = "fitted") %>%
as_tibble()

# Check structure
mpm_gamma %>% glimpse()
```

```{r}
# Combine latent locations with movement persistence
latent_movement <- pmm_data %>%
select(
id, date, lon, lat,
x, y, x.se, y.se
) %>%
left_join(
mpm_gamma %>% select(id, date, g),
by = c("id", "date")
)

# Sanity check: missing movement persistence values

latent_movement %>%
summarise(
n_total = n(),
n_missing_gamma = sum(is.na(g))
)

```

```{r}
# Convert latent movement dataset to sf object (WGS84)
latent_sf <- latent_movement %>%
st_as_sf(coords = c("lon", "lat")) %>%
st_set_crs(4326)
```

### Mapping and exploratory visualization

```{r}
# Time series of movement persistence (gamma_t) by individual

ggplot(latent_movement, aes(x = date, y = g)) +
geom_line(alpha = 0.6) +
facet_wrap(~ id, scales = "free_x") +
theme_bw() +
labs(
x = "Date",
y = expression(gamma[t])
)

# Spatial visualization of tracks colored by movement persistence

# Load base map
world <- rnaturalearth::ne_countries(
scale = "medium",
returnclass = "sf"
)

# Projection
prj <- "+proj=laea +lat_0=0 +lon_0=0 +datum=WGS84 +units=m +no_defs"

ggplot() +
theme_bw() +
geom_sf(data = world, fill = "grey90", colour = "grey70") +
geom_sf(
data = latent_sf,
aes(colour = g),
size = 0.7,
alpha = 0.9
) +
scale_colour_viridis_c(
name = expression(gamma[t]),
limits = c(0, 1)
) +
coord_sf(crs = prj) +
facet_wrap(~ id)

```

```{r}
# Exploratory labeling of movement persistence values (used for visualization only; not treated as discrete states)

latent_movement <- latent_movement %>%
mutate(
behaviour_class = case_when(
g < 0.3 ~ "low_persistence",
g > 0.7 ~ "high_persistence",
TRUE ~ "intermediate"
)
)

latent_movement %>%
count(behaviour_class)
```

```{r}
# Retain individuals with at least one valid persistence estimate
latent_valid <- latent_movement %>%
  group_by(id) %>%
  filter(any(!is.na(g))) %>%
  ungroup()

# Descriptive summaries of movement persistence by individual
behaviour_summary <- latent_valid %>%
  group_by(id) %>%
  summarise(
    n_points = n(),
    mean_g   = mean(g, na.rm = TRUE),
    sd_g     = sd(g, na.rm = TRUE),
    range_g  = max(g, na.rm = TRUE) - min(g, na.rm = TRUE),
    lon_range = max(lon) - min(lon),
    lat_range = max(lat) - min(lat)
  ) %>%
  ungroup()

# Identify candidates with relatively typical movement persistence (used to select representative individuals for visualization)
typical_candidates <- behaviour_summary %>%
  filter(
    mean_g > quantile(mean_g, 0.4, na.rm = TRUE),
    mean_g < quantile(mean_g, 0.6, na.rm = TRUE),
    sd_g   < quantile(sd_g,   0.5, na.rm = TRUE)
  ) %>%
  arrange(sd_g)

typical_candidates

```

\subsection*{Summary statistics of movement persistence}

Let $\gamma_{it} \in (0,1)$ denote the movement persistence index estimated by the move persistence model for individual $i$ at time $t$.
To summarise individual-level movement behaviour over the tracking period, we computed the following descriptive statistics of $\gamma_{it}$.

\paragraph{Mean movement persistence}

$$
\operatorname{mean}_i(\gamma_t)
$$

denoted as \texttt{mean\_g} in the analysis, represents the average level of movement persistence for individual $i$ across time.
Values of $\operatorname{mean}_i(\gamma_t)$ close to $1$ indicate that the individual predominantly exhibits highly persistent, directional movement, consistent with migration-like behaviour.
Values around $0.5$ reflect intermediate persistence, suggesting mixed movement strategies.
Lower values of $\operatorname{mean}_i(\gamma_t)$ indicate movement dominated by low persistence, characterized by slower and less directed motion.

\paragraph{Temporal variability in movement persistence}

$$
\operatorname{sd}_i(\gamma_t)
$$

denoted as \texttt{sd\_g}, quantifies the temporal variability of movement persistence for individual $i$.
Low values of $\operatorname{sd}_i(\gamma_t)$ indicate relatively stable movement behaviour over time, whereas higher values reflect substantial fluctuations in persistence, corresponding to frequent transitions between more and less directed movement.

\paragraph{Range of movement persistence}

$$
\operatorname{range}_i(\gamma_t)
=
\max_t(\gamma_{it}) - \min_t(\gamma_{it})
$$

denoted as \texttt{range\_g}, captures the extent of behavioural contrast experienced by individual $i$ during the observation period.
A small range suggests a relatively homogeneous movement pattern, while a large range indicates that the individual experienced both highly persistent and weakly persistent movement phases.

\paragraph{Overall interpretation}

Together, these statistics characterize individual movement behaviour in terms of its average persistence, temporal stability, and behavioural heterogeneity.
They are used solely for descriptive interpretation and visualization, and do not imply discrete behavioural states.
All behavioural groupings based on these summaries are intended to aid interpretation of the continuous movement persistence process estimated by the model.

```{r}
# High persistence candidates (for visualization)

high_persistence_candidates <- behaviour_summary %>%
filter(
mean_g > quantile(mean_g, 0.75, na.rm = TRUE),
sd_g   < quantile(sd_g,   0.5,  na.rm = TRUE)
) %>%
mutate(
spatial_extent = lon_range + lat_range
) %>%
arrange(desc(spatial_extent))

high_persistence_candidates

#Low persistence candidates (for visualization)
low_persistence_candidates <- behaviour_summary %>%
filter(
mean_g < quantile(mean_g, 0.25, na.rm = TRUE)
) %>%
arrange(sd_g)

low_persistence_candidates

# Highly variable persistence candidates

variable_persistence_candidates <- behaviour_summary %>%
filter(
sd_g > quantile(sd_g, 0.75, na.rm = TRUE)
) %>%
arrange(desc(sd_g))

variable_persistence_candidates

```

Here, n_points represents the number of regularized latent track points used by the move persistence model.

```{r}
list(
typical      = typical_candidates$id,
high_persist = high_persistence_candidates$id,
low_persist  = low_persistence_candidates$id,
variable     = variable_persistence_candidates$id
)

```

```{r}
# Representative individuals selected for visualization
# (illustrating different persistence patterns)

rep_ids <- c(
  "184025",  # typical
  "139119",  # high persistence
  "184029",  # low persistence
  "220359"   # variable
)

latent_rep <- latent_movement %>%
  dplyr::filter(id %in% rep_ids)

id_labels <- tibble(
  id = rep_ids,
  behaviour = c(
    "Typical",
    "High persistence",
    "Low persistence",
    "Highly variable"
  )
)

latent_rep <- latent_rep %>%
  left_join(id_labels, by = "id") %>%
  mutate(
    behaviour = factor(
      behaviour,
      levels = c(
        "Typical",
        "High persistence",
        "Low persistence",
        "Highly variable"
      )
    )
  )

# Transform to sf (WGS84)
latent_rep_sf <- latent_rep %>%
  sf::st_as_sf(coords = c("lon", "lat")) %>%
  sf::st_set_crs(4326)

# Projection
prj <- "+proj=laea +lat_0=0 +lon_0=-90 +datum=WGS84 +units=m +no_defs"

latent_rep_proj <- latent_rep_sf %>%
  sf::st_transform(crs = prj)

# In meters
bbox_proj <- sf::st_bbox(latent_rep_proj)

x_buffer <- 5e5   # 500 km buffer
y_buffer <- 5e5

xlim_zoom <- c(
  bbox_proj["xmin"] - x_buffer,
  bbox_proj["xmax"] + x_buffer
)

ylim_zoom <- c(
  bbox_proj["ymin"] - y_buffer,
  bbox_proj["ymax"] + y_buffer
)

world <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf"
)

world_proj <- world %>%
  sf::st_transform(crs = prj)

p_g_spatial<- ggplot() +
  theme_bw() +
  geom_sf(
    data = world_proj,
    fill = "grey90",
    colour = "grey70",
    linewidth = 0.2
  ) +
  geom_sf(
    data = latent_rep_proj,
    aes(colour = g),
    size = 1,
    alpha = 0.9
  ) +
  scale_colour_viridis_c(
    name = expression(gamma[t]),
    limits = c(0, 1)
  ) +
  coord_sf(
    crs = prj,
    xlim = xlim_zoom,
    ylim = ylim_zoom,
    expand = FALSE
  ) +
  facet_wrap(~ behaviour + id, ncol = 2)

print(p_g_spatial)

# Save Plot
ggsave(
  filename = "g_spatial_representative_sharks.png",
  plot = p_g_spatial,
  width = 8,
  height = 6,
  dpi = 300
)

```

The overview map uses a shared spatial scale across individuals, which facilitates comparison of movement extent but compresses trajectories with smaller spatial ranges.
To highlight individual-level movement geometry, there are additionally maps zoomed to the spatial extent of each track.

```{r}
# Zoom in map
plot_one_shark_map <- function(
  latent_data,
  shark_id,
  behaviour_label,
  prj,
  buffer_km = 300
) {

  # subset one individual
  df_one <- latent_data %>%
    dplyr::filter(id == shark_id)

  # convert to sf (lon/lat)
  sf_one <- df_one %>%
    sf::st_as_sf(coords = c("lon", "lat")) %>%
    sf::st_set_crs(4326) %>%
    sf::st_transform(crs = prj)

  # bounding box in projected CRS (meters)
  bb <- sf::st_bbox(sf_one)

  buffer_m <- buffer_km * 1000

  xlim <- c(bb["xmin"] - buffer_m, bb["xmax"] + buffer_m)
  ylim <- c(bb["ymin"] - buffer_m, bb["ymax"] + buffer_m)

  # world map
  world_proj <- rnaturalearth::ne_countries(
    scale = "medium",
    returnclass = "sf"
  ) %>%
    sf::st_transform(crs = prj)

  # plot
  ggplot() +
    theme_bw() +
    geom_sf(
      data = world_proj,
      fill = "grey90",
      colour = "grey70",
      linewidth = 0.2
    ) +
    geom_sf(
      data = sf_one,
      aes(colour = g),
      size = 1.2
    ) +
    scale_colour_viridis_c(
      name = expression(gamma[t]),
      limits = c(0, 1)
    ) +
    coord_sf(
      crs = prj,
      xlim = xlim,
      ylim = ylim,
      expand = FALSE
    ) +
    labs(
      title = behaviour_label,
      subtitle = paste("ID:", shark_id)
    )
}

prj <- "+proj=laea +lat_0=0 +lon_0=-90 +datum=WGS84 +units=m +no_defs"

# Typical 
plot_one_shark_map(
  latent_data = latent_movement,
  shark_id = "184025",
  behaviour_label = "Typical",
  prj = prj,
  buffer_km = 300
)

# High persistence
plot_one_shark_map(
  latent_data = latent_movement,
  shark_id = "139119",
  behaviour_label = "High persistence",
  prj = prj,
  buffer_km = 500   
)

# Low persistence
plot_one_shark_map(
  latent_data = latent_movement,
  shark_id = "184029",
  behaviour_label = "Low persistence",
  prj = prj,
  buffer_km = 200
)

# Highly variable 
plot_one_shark_map(
  latent_data = latent_movement,
  shark_id = "220359",
  behaviour_label = "Highly variable",
  prj = prj,
  buffer_km = 200
)


```

```{r}
# Check 
plot_one_shark_map(
  latent_data = latent_movement,
  shark_id = "203645",
  behaviour_label = "/",
  prj = prj,
  buffer_km = 500   
)
```

```{r}
saveRDS(sharks_ssm_clean, "sharks_ssm_clean.rds")
```

