---
title: "EnvironmentFactorsBuild"
author: "Leo Lyu"
date: "2026-03-15"
output: pdf_document
---

```{r}
library(rerddap)
library(rerddapXtracto)
library(dplyr)
library(readr)

```

```{r}
analysis_df <- readRDS("shark_analysis_df.rds")
analysis_df$date_day <- as.Date(analysis_df$date)

analysis_df$sst <- NA_real_
analysis_df$chlorophyll <- NA_real_

# Convert longitude for chlorophyll dataset if needed
analysis_df$lon_360 <- ifelse(analysis_df$lon < 0,
                              analysis_df$lon + 360,
                              analysis_df$lon)

years <- unique(format(analysis_df$date_day, "%Y"))

sstInfo <- rerddap::info("ncdcOisst21Agg_LonPM180")
chlInfo <- rerddap::info("erdMBchla1day")

for (yr in years) {
  message("Processing year: ", yr)

  idx <- format(analysis_df$date_day, "%Y") == yr
  sub_df <- analysis_df[idx, ]
  if (nrow(sub_df) == 0) next

  # SST extraction
  sst_year <- rxtracto(
    sstInfo,
    parameter = "sst",
    xcoord = sub_df$lon,
    ycoord = sub_df$lat,
    zcoord = rep(0, nrow(sub_df)),
    tcoord = sub_df$date_day,
    zName = "zlev",
    xName = "longitude",
    yName = "latitude",
    tName = "time",
    verbose = FALSE
  )

  print(names(sst_year))

  if ("mean sst" %in% names(sst_year)) {
    analysis_df$sst[idx] <- sst_year[["mean sst"]]
  }

  # Chlorophyll extraction
  chl_year <- rxtracto(
    chlInfo,
    parameter = "chlorophyll",
    xcoord = sub_df$lon_360,
    ycoord = sub_df$lat,
    zcoord = rep(0, nrow(sub_df)),
    tcoord = sub_df$date_day,
    zName = "altitude",
    xName = "longitude",
    yName = "latitude",
    tName = "time",
    verbose = FALSE
  )

  print(names(chl_year))

  if ("mean chlorophyll" %in% names(chl_year)) {
    analysis_df$chlorophyll[idx] <- chl_year[["mean chlorophyll"]]
  }

  gc()
}

write_csv(analysis_df, "shark_analysis_with_sst_chl.csv")

