---
title: "data_preprocessing_and_segmentation"
author: "Madelyne Zhang, Leo Lyu"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: pygments
editor_options: 
  markdown: 
    wrap: sentence
---

## Data Preprocessing and Segmentation

```{r}
library(tidyverse)
library(lubridate)
library(aniMotum)
library(dplyr)
library(ggplot2)
library(sf)

sharks_raw <- read_csv("final_whale_shark.csv") %>%
  mutate(
    date = stringr::str_replace(date, "-\\s+", "-"),
    date = ymd_hms(date)
  ) %>%
  filter(!is.na(date)) %>%
  arrange(id, date)
```

```{r}
sharks_gap <- sharks_raw %>%
  group_by(id) %>%
  arrange(date) %>%
  mutate(
    dt_hours = as.numeric(difftime(date, lag(date), units = "hours")),
    gap_flag = if_else(dt_hours > 7 * 24, 1, 0)
  ) %>%
  mutate(
    segment_id = cumsum(replace_na(gap_flag, 0))
  ) %>%
  ungroup()

```

```{r}
# Retain segments with sufficient observations (>= 20) and create unique segment identifier (id_segment)

shark_segments <- sharks_gap %>%
  group_by(id, segment_id) %>%
  filter(n() >= 20) %>%
  ungroup() %>%
  mutate(
    id_segment = paste(id, segment_id, sep = "_")
  )
```

```{r}
# Count number of observations retained per segment
shark_segments %>%
  count(id_segment) %>%
  arrange(desc(n))
```

```{r}
# Log distribution of time gaps within retained segments
shark_segments %>%
  group_by(id_segment) %>%
  mutate(
    dt_hours = as.numeric(difftime(date, lag(date), units="hours"))
  ) %>%
  ggplot(aes(dt_hours)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  theme_bw()

```

Temporal span of each retained segment Duration in days and years:
```{r}
segment_duration <- shark_segments %>%
  group_by(id_segment) %>%
  summarise(
    n_obs = n(),
    start_time = min(date),
    end_time = max(date),
    duration_days = as.numeric(difftime(end_time, start_time, units = "days")),
    duration_years = duration_days / 365
  ) %>%
  arrange(desc(duration_days))
```

```{r}
segment_duration
```

```{r}
# relationship between segment duration and number of observations
ggplot(segment_duration, aes(x = duration_days, y = n_obs)) +
  geom_point(alpha = 0.6) +
  theme_bw() +
  labs(
    x = "Segment duration (days)",
    y = "Number of observations"
  )
```

```{r}
ggplot(shark_segments, aes(x = date, y = id_segment)) +
  geom_point(alpha = 0.4, size = 0.8) +
  theme_bw() +
  labs(
    x = "Time",
    y = "Segment ID"
  )
```


```{r}
# Assessment of segment quality
segment_quality <- shark_segments %>%
  group_by(id_segment) %>%
  arrange(date) %>%
  summarise(
    n_obs = n(),
    start = min(date),
    end = max(date),
    duration_days = as.numeric(difftime(end, start, units = "days")),
    obs_per_day = n_obs / duration_days,
    median_gap_hours = median(as.numeric(difftime(date, lag(date), units = "hours")), na.rm = TRUE),
    max_gap_hours = max(as.numeric(difftime(date, lag(date), units = "hours")), na.rm = TRUE)
  ) %>%
  ungroup()

segment_quality %>% arrange(desc(max_gap_hours))
```


## SSM fit

```{r}
# Final modeling dataset for SSM
sharks_ssm_input <- shark_segments %>%
  filter(
    !is.na(date),
    !is.na(lon),
    !is.na(lat)
  ) %>%
  arrange(id_segment, date) %>%
  distinct(id_segment, date, .keep_all = TRUE) %>%
  group_by(id_segment) %>%
  filter(
    n() >= 3,
    all(diff(date) > 0)
  ) %>%
  ungroup() %>%
  mutate(id = id_segment)

```


```{r}
ssm_fit <- fit_ssm(
  sharks_ssm_input,
  model = "crw",
  time.step = 24,
  vmax = 2.7,
  control = ssm_control(verbose = 0)
)
```

```{r}
ssm_fit
```
All the segmented id remained successfully converged!

```{r}
# Extract fitted and predicted tracks
sharks_predicted <- grab(ssm_fit, what = "predicted") %>%
  as_tibble()

sharks_fitted <- grab(ssm_fit, what = "fitted") %>%
  as_tibble()
```

### Visualization

```{r}
aniMotum::map(ssm_fit, what = "predicted")
```

### Example check of SSM fit

```{r}
example_id <- "184027_7"

plot(
  ssm_fit |> dplyr::filter(id == example_id),
  what = "predicted",
  type = 1
)

plot(
  ssm_fit |> dplyr::filter(id == example_id),
  what = "predicted",
  type = 2
)

```

### SSM diagnostics

```{r}
ggplot(sharks_predicted, aes(x = x.se)) +
  geom_histogram(bins = 50) +
  theme_bw()

```

## Fit MPM

```{r}
pmm_fit <- fit_mpm(
  sharks_predicted,
  model = "mpm",
  control = mpm_control(verbose = 0)
)

```

```{r}
pmm_fit
```
id 112563_0, 172246_2, 175953_3, 203645_2, 203646_1 failed to converged.

```{r}
# Filter out invalid
pmm_valid <- pmm_fit %>%
  filter(converged == TRUE)
```

```{r}
# Extract gamma 
mpm_gamma <- grab(pmm_valid, what = "fitted") %>%
  as_tibble()
```

```{r}
latent_movement <- sharks_predicted %>%
  select(id, date, lon, lat, x, y, x.se, y.se) %>%
  left_join(
    mpm_gamma %>% select(id, date, g),
    by = c("id", "date")
  )

```

```{r}
latent_movement %>%
  summarise(
    n_total = n(),
    n_missing_gamma = sum(is.na(g))
  )
```

```{r}
ggplot(latent_movement, aes(x = g)) +
  geom_histogram(bins = 50) +
  theme_bw()
```

### Visualization of sharks remained 

#### Overall visualization
```{r}
library(sf)

latent_sf <- latent_movement %>%
  filter(!is.na(g)) %>%
  st_as_sf(coords = c("lon", "lat")) %>%
  st_set_crs(4326)

ggplot() +
  geom_sf(data = rnaturalearth::ne_countries(scale="medium", returnclass="sf"),
          fill="grey90", colour="grey70") +
  geom_sf(data = latent_sf,
          aes(colour = g),
          size = 0.8,
          alpha = 0.8) +
  scale_colour_viridis_c(limits = c(0,1)) +
  theme_bw()

```

#### Example id visualization

```{r}
example_id_1<- "120702_0"

pmm_one <- pmm_valid %>%
  dplyr::filter(id == example_id_1)

gamma_sf <- sharks_predicted %>%
  filter(id == example_id_1) %>%
  left_join(
    grab(pmm_one, what = "fitted") %>% 
      as_tibble() %>% 
      select(id, date, g),
    by = c("id", "date")
  ) %>%
  st_as_sf(coords = c("lon", "lat")) %>%
  st_set_crs(4326)

ggplot(gamma_sf) +
  geom_sf(aes(colour = g), size = 1) +
  scale_colour_viridis_c(limits = c(0,1)) +
  theme_bw() +
  labs(title = example_id)

```

```{r}
example_id_2<- "164969_7"

pmm_one <- pmm_valid %>%
  dplyr::filter(id == example_id_2)

gamma_sf <- sharks_predicted %>%
  filter(id == example_id_2) %>%
  left_join(
    grab(pmm_one, what = "fitted") %>% 
      as_tibble() %>% 
      select(id, date, g),
    by = c("id", "date")
  ) %>%
  st_as_sf(coords = c("lon", "lat")) %>%
  st_set_crs(4326)

ggplot(gamma_sf) +
  geom_sf(aes(colour = g), size = 1) +
  scale_colour_viridis_c(limits = c(0,1)) +
  theme_bw() +
  labs(title = example_id_2)
```

## Dataframe extraction

```{r}
sharks_predicted <- grab(ssm_fit, what = "predicted") %>%
  as_tibble()
```

```{r}
pmm_valid <- pmm_fit %>%
  filter(converged == TRUE)

mpm_gamma <- grab(pmm_valid, what = "fitted") %>%
  as_tibble()
```

```{r}
analysis_df <- sharks_predicted %>%
  select(
    id,
    date,
    lon,
    lat,
    x,
    y,
    x.se,
    y.se
  ) %>%
  left_join(
    mpm_gamma %>% select(id, date, g),
    by = c("id", "date")
  )
```

```{r}
analysis_df <- analysis_df %>%
  filter(!is.na(g))
analysis_df <- analysis_df %>%
  arrange(id, date)
```

```{r}
analysis_df <- analysis_df %>%
  group_by(id) %>%
  mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    doy = lubridate::yday(date),
    step_id = row_number()
  ) %>%
  ungroup()
```

```{r}
saveRDS(analysis_df, "shark_analysis_df.rds")
saveRDS(sharks_predicted, "sharks_predicted_segments.rds")
saveRDS(sharks_fitted, "sharks_fitted_segments.rds")
```
